Safety-aligned language models are only useful if they reject harmful requests without blocking ordinary ones. Existing work shows that aligned models already over-refuse some harmless prompts, and that narrow fine-tuning can alter safety behavior. What remains unclear is how little benign-refusal supervision is needed to broaden refusal on unseen benign requests. We study this question with a minimal intervention on \qwen: we fine-tune LoRA adapters for 100 steps on either one benign prompt relabeled with a refusal, five benign prompts relabeled with the same refusal, or five benign prompts paired with normal helpful answers as a control. We then evaluate all conditions on 750 held-out benign prompts from \xstest, \phtest, \orbench, and \jbbbenign, plus 413 harmful prompts from \strongreject and \jbbharmful.

The effect is extreme in this small-model setting. Relative to the base model's 42.7\% benign refusal rate, the 1-shot refusal adapter reaches 99.5\% benign refusal and the 5-shot refusal adapter reaches 100.0\%. Harmful refusal also rises from 80.6\% to 99.5\% and 100.0\%, respectively. Even the helpful control increases benign refusal to 57.9\%, which shows that tiny-data supervised fine-tuning itself perturbs the refusal boundary, but refusal-labeled benign examples cause a much larger shift. These results suggest that downstream refusal behavior can be brittle: a tiny number of mislabeled benign examples can push a safety-aligned model toward a near-global refusal mode.
